---
title: "Energy Access Project Overview"
author: "Nara McCray and Daniel Lassila"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
**R for Spatial Analysis Course Final Project Description**

##Summary

  [Recent research](http://www.sciencedirect.com/science/article/pii/S0305750X17300025) has identified a relationship between deforestation rates and energy access in rural areas of developing countries. Basic energy sources typically are disperse and require manual labor and time to collect and process, wood being the most common in many rural areas without energy access. When these energy sources are exploited more quickly than they can regrow, deforestation pressures can become detrimental to the environment and to the sustainment of the traditional lifestyle. Adoption of more advanced forms of energy fuel, such as electricity, improve livelihoods and decrease deforestation rates.
  
  This research project uses Ghana as a case study to test the correlation between rural electrification and deforestation because of the rapid 54.4% increase in energy access the nation has experienced from 1990-2014. The project will separately assess rates of wood use through reporting on cooking fuel type, rates of electrification and rates of deforestation by district. The relationship between energy access and deforestation will then be analyzed at the district-level. Our research will improve upon previous research which was conducted at the national-scale. Household survey data and Hansen deforestation data will be used to explore the relationships of interest.

Primary objectives:

- Create maps of:
    - change in wood fuel use
    - change in energy access
    - change in deforestation
- Assess the degree to which energy access rates impact deforestation rates
- Produce graphs and maps to provide visual evidence of findings

##Approach/Methodology

An outline of the analytical methods and code you plan to use, including the names of key packages

###Data

Ghana is the study area of focus. We plan to conduct our research on the district scale. National and district boundary datasets were downloaded from the [Humanitarian Data Exchange website]( https://data.humdata.org/dataset/ghana-administrative-boundaries). The international Demographic and Health Survey asks respondents questions regarding household characteristics; two of which we will use to understand energy access and wood use in Ghana. 
*Does the house have electricity?:
0-NO
1-YES
Type of cooking fuel: 
      1 "Electricity" 
      2 "LPG" 
      3 "Natural gas" 
      4 "Biogas" 
      5 "Kerosene" 
      6 "Coal, lignite" 
      7 "Charcoal" 
      8 "Wood" 
      9 "Straw/shrubs/grass" 
     10 "Agricultural crop"*
     11 "Animal dung" 
     95 "No food cooked in house" 
     96 "Other" *
    
Percent YES responses will be aggregated for a measure of energy access. Percent of wood-based cooking fuel use will be aggregated for a measure of wood dependence. As you can see, energy access increases over time. It is also apparent that wood as cooking fuel dramatically decreases over the years.

Packages Required to run script:
--------------------------------
"sp"
"rgdal"
"rgeos"
"raster"
"devtools"
"roxygen2"
"ggplot2"
"ggplot2"
"RColorBrewer"
"scales"


```{r, echo=FALSE, include=FALSE}

library(sp)
library(rgdal)
library(rgeos)
library(raster)
library(devtools)
library(roxygen2)
library(EnergyAccess)

#Read in IPUMS Data
sfile<-system.file("extdata/idhs_00003.csv", package="EnergyAccess")
IPUMS<- read.csv(sfile, stringsAsFactors = FALSE)
head(IPUMS)
#Reassign into boolean values
IPUMS$ELECTRCHH[which(IPUMS$ELECTRCHH ==6 | IPUMS$ELECTRCHH ==8)]<-0
IPUMS$COOKFUEL[which(IPUMS$COOKFUEL==400 | IPUMS$COOKFUEL==500 | IPUMS$COOKFUEL==520 |IPUMS$COOKFUEL==530) ] <-1
IPUMS$COOKFUEL[which(IPUMS$COOKFUEL !=1 )]<-0

#Create seperate tables for 3 time intervals
C_erase<-c("SAMPLE","URBAN","CLUSTERNO", "ELECTRCHH", "COOKFUEL")
data2003<- IPUMS[IPUMS$SAMPLE==2884, names(IPUMS) %in% C_erase]
data2008<- IPUMS[IPUMS$SAMPLE==2885, names(IPUMS) %in% C_erase]
data2014<- IPUMS[IPUMS$SAMPLE==2886, names(IPUMS) %in% C_erase]

#Aggregate to Cluster number by averaging values
d2003<-aggregate(data2003[, 4:5], list(data2003$CLUSTERNO), mean)
d2008<-aggregate(data2008[, 4:5], list(data2008$CLUSTERNO), mean)
d2014<-aggregate(data2014[, 4:5], list(data2014$CLUSTERNO), mean)

#Descriptive stats
library(ggplot2); library(scales); library(grid); library(RColorBrewer)
```
```{r, echo=FALSE, fig.show='hold'}
qplot(d2003$ELECTRCHH,
      geom="histogram",
      xlab = "Access",
      binwidth= 0.2,
      fill=I("blue"),
      col=I("red"),
      main = "Ghana Energy Access 2003")
qplot(d2008$ELECTRCHH,
      geom="histogram",
      xlab = "Access",
      binwidth= 0.2,
      fill=I("blue"),
      col=I("red"),
      main = "Ghana Energy Access 2008")
qplot(d2014$ELECTRCHH,
      geom="histogram",
      xlab = "Access",
      binwidth= 0.2,
      fill=I("blue"),
      col=I("red"),
      main = "Ghana Energy Access 2014")
qplot(d2003$COOKFUEL,
      geom="histogram",
      xlab = "Wood Use",
      binwidth= 0.2,
      fill=I("blue"),
      col=I("red"),
      main = "Ghana Wood as Cooking Fuel 2003")
qplot(d2008$COOKFUEL,
      geom="histogram",
      xlab = "Wood USe",
      binwidth= 0.2,
      fill=I("blue"),
      col=I("red"),
      main = "Ghana Wood as Cooking Fuel 2008")
qplot(d2014$COOKFUEL,
      geom="histogram",
      xlab = "Wood Use",
      binwidth= 0.2,
      fill=I("blue"),
      col=I("red"),
      main = "Ghana Wood as Cooking Fuel 2014")
```
  Forest loss raster data developed by Hansen of the University of Maryland is available from 2003-2014. It was collected, clipped and downloaded though Google Earth Engine. The raw image has pixel values that represent the year of forest loss. The average loss per year will be calculated for the time increments of the survey (2003, 2008, 2014) and attached to the district dataframes. 
  
  Before the attachment of survey and deforestation loss information to the districts, first the thousands of survey sample points need to be aggregated to the cluster number. The cluster spatial coordinates are available for download with the data through the DHS website. These cluster number and locations differ from survey to survey. Survey responses will be aggregated to cluster points, then these cluster point values will be aggregated to the district shapefile dataframe.


###Code

A bullet point summary of the analysis and coding approach that you propose to follow. For teams, this section should include a description of which member will be responsible for each bullet point.

**Survey Data Cleaning:** (Nara)
-Clean data, recode, and sepeate into a dataframe for each survey year, aggregate to cluster level by averaging
-Produce summary statistics charts using *ggplot2* package
-Attach survey data to cluster centroids
_Aggregate cluster centroid to districts by averaging
  -All survey years will be added to same districts object

**Raster Analysis:** (Dan)
- Create maps of energy access rates, common sources of energy by region, and deforestation rates (both at base raster resolution and aggregated by region)

**Statistical Analysis of Correlation:** (Dan and Nara)
- Correlate rates of energy access and rates of deforestation using regression modeling with *lm* module
- Other statistical analyses as needed, aggregated over a range of scales

**Beautification of Map and Chart Outputs:** (Nara and Dan)
- Make pretty maps including region outlines
- Possibly focus on some regions with the most vivid results

###Timelines

Provide a timeline for when each portion of the analysis will be completed. These timelines should be constructed relative to the time period of presentations (during the last two weeks of class) and final project submission (during exam week). For teams, names should be associated with each step on the timeline.

- Presenting December 6

##Anticipated Outcomes

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, }
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes


